Altering Backward Pass Gradients improves Convergence

### 2022-09-19 paper.tex
Mod7

\begin{figure}[!t]
\centering
\begin{subfigure}{.5\columnwidth}
\centering
\includegraphics[width=0.9\columnwidth]{acc_vs_epoch_r18}
\caption{Plot of training and test accuracies (ResNet-18)}
\end{subfigure}%
\begin{subfigure}{.5\columnwidth}
\centering
\includegraphics[width=0.9\columnwidth]{acc_vs_epoch_r50}
\caption{Plot of training and test accuracies (ResNet-50)}
\end{subfigure}
\caption{ Log-linear plots of training and test accuracies and comparison with baseline
of batch-normalized variants: \textbf{(a)} ResNet-18 ($\alpha=0.25$), \textbf{(b)}
ResNet-50 ($\alpha=0.3$). For better visualization of accuracies obtained in the final
iterations, we plot the y-axis in log scale. PowerGrad Transform improves upon the
baseline in all cases. The corresponding accuracies are mentioned in table
\ref{tab:imagenet_table}. }
\label{fig:metrics}
\end{figure}

@@

In our experiments with Squeeze-and-Excitation variant of ResNet-50 i.e.
SE-ResNet-50\cite{hu2018squeeze}, we observe significant improvements; a $0.97\%$ boost
in training accuracy and a $0.734\%$ increase in test accuracy. These improvements are
obtained with the number of epochs being kept the same as that of the baseline model.
Alternatively, practioners can choose to employ PGT and have the networks converge to
accuracies similar to that of the baseline with a much lower epoch budget. We explore
this in detail in section \ref{sec:speed}. Therefore, PGT also speeds up training. In
our experiments with ResNet-50, we find an increase of $0.5\%$ and $0.656\%$ performance
enhancement over the baseline while using the cosine scheduler. The corresponding
improvement over the step scheduler baseline for ResNet-50 is $0.57\%$ and $0.524\%$
respectively for training and testing. ResNet-101 sees a higher improvement in training
fit, $0.81\%$ to be exact, while the improvement over the test set is $0.362\%$. Smaller
networks such as SE-ResNet-18 and ResNet-18 sees accuracy boosts which are smaller but
nevertheless positive. Therefore we conclude that with PGT, the consistent improvements
in training accuracies across all cases is because the networks train better and arrive
at better optimas during convergence. Per epoch training and test accuracy plots of
ResNet-18 and ResNet-50 (both with and without PGT) are shown in Fig.
\ref{fig:metrics}(a, b). For better visualization of the top end of the plots, we scale
the y-axis in log scale. Our experiments with batch-normalized networks indicate that
PowerGrad Transform speeds up training and allows the network to learn better
representations\footnote{Reproducible code, training recipes for the experiments,
pretrained checkpoints and training logs are provided at: \\
\url{ https://github.com/bishshoy/power-grad-transform }. Code is adapted from
\cite{rw2019timm}.}.

@@

\begin{table}[!t]
\centering
\caption{ Results for non-normalized ResNets on ImageNet-1K. Each experiment is
conducted with a 100 epoch budget using the cosine decay scheduler. Best training and
test accuracies are highlighted in red and blue respectively. Top differences in
training and test accuracies are marked in yellow. The baselines in each segment is
taken as the reference point for all `Diff' accuracies in that segment. The first three
segments highlight comparison for experiments where the methods are applied
individually, while the last segment denotes performance metrics for combinations of
different methods (GC + PGT, AGC + PGT). We see that at any batch size, with PGT, the
performance of non-normalized version of ResNet-18 improves significantly over the
baseline owing to the improvements in training as observed in Fig. \ref{fig:agc_pgt}. }
\label{tab:wobn_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Batch Size}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PowerGrad} & \textbf{\hspace{-0.65cm} Train}
& \textbf{Train} & \textbf{\hspace{-0.65cm} Test} & \textbf{Test} \\
& & & \textbf{Transform ($\alpha$)} & \textbf{Accuracy (\%)} & \textbf{Diff (\%)} &
\textbf{Accuracy (\%)} & \textbf{Diff (\%)} \\
\midrule
ResNet-18 (non-BN) & 1024 & Baseline & - & 67.19 & - & 64.93 & - \\
\cmidrule{2-8}
& 512 & Baseline & - & 68.02 & - & 66.552 & - \\
& 512 & PGT & 0.25 & \textcolor{red}{\textbf{69.5}} & \textcolor{olive}{\textbf{+1.48}} &
\textcolor{blue}{\textbf{67.236}} & \textcolor{olive}{\textbf{+0.684}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC & - & 69.04 & +0.18 & 67.064 & +0.268 \\
& 256 & AGC & - & 69.06 & +0.2 & 67.298 & +0.502 \\
& 256 & PGT & 0.25 & \textcolor{red}{\textbf{69.97}} & \textcolor{olive}{\textbf{+1.11}} &
\textcolor{blue}{\textbf{67.814}} & \textcolor{olive}{\textbf{+1.018}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC+PGT & 0.25 & 68.67 & -0.19 & 67.088 & +0.292 \\
& 256 & AGC+PGT & 0.25 & \textcolor{red}{\textbf{70.92}} & \textcolor{olive}{\textbf{+2.06}} &
\textcolor{blue}{\textbf{68.856}} & \textcolor{olive}{\textbf{+2.06}} \\
\end{tabular}
\end{table}

We provide accuracy results for different batch sizes for the non-BN ResNet-18 trained
on ImageNet-1K and also experiment with different methods (GC, AGC, PGT along with their
combinations). Table \ref{tab:wobn_table} shows the findings. To begin, we use a batch
size of 1024. This is only to illustrate that in unnormalized networks, accuracy suffers
significantly when batch sizes are large. Additionally, the training process is often
unstable at this batch size, resulting in frequent layer dropouts (as seen in
\ref{sec:Empi}) and infinite loss values. Thus for all subsequent tests, we use batch
sizes of 512 and 256. At a batch size of 512, invoking PGT with $\alpha=0.25$ improves
the training accuracy baseline by $1.48\%$ and the test accuracy baseline by $0.684\%$,
while at a batch size of 256, the improvement in training and test accuracies are
$1.11\%$ and $1.018\%$ respectively. In comparison, the test accuracy improvements
obtained by GC and AGC at batch size of 256 is much less, at $0.27\%$ and $0.5\%$,
respectively.

@@

We conduct an ablation study to investigate the effects of PowerGrad Transform for
different values of the PGT hyperparameter($\alpha$). Grid plots are made available in
the supplementary section. We have previously demonstrated from the experiments depicted
in section \ref{sec:Expe}, that PowerGrad Transform causes the network to arrive at a
better optima and fit the training dataset better than the baseline. In this ablation
study, we use the ResNet-50 architecture and we combine our proposed method with other
regularization techniques such as label smoothing and report our findings in table
\ref{tab:ablation_table}. We also investigate the two optimal peaks of the
hyperparameter ($\alpha$) obtained from ResNet-50's grid search ($\alpha=0.05$ and
$\alpha=0.3$). First, we examine the effect of PGT on the step scheduler baseline in
order to later compare it to the cosine scheduler baseline. PGT improves upon the step
scheduler baseline (test set) by a substantial margin with $0.524\%$ ($76.494\%$ as
opposed to $75.97\%$). Introducing the cosine scheduler yields a $0.59\%$ improvement
($76.56\%$ vs. $75.97\%$) over the step scheduler. After introducing label smoothing,
the test accuracy relative to the cosine scheduler baseline increases by only $0.138\%$
(from $76.56\%$ to $76.698\%$). However, introducing PGT with $\alpha=0.3$ alone
(without label smoothing) improves the cosine scheduler baseline by $0.326\%$
($76.886\%$ vs. $76.56\%$). Combining PGT ($\alpha=0.3$) with label smoothing improves
the performance on the test set further by $0.408\%$ (from $76.56\%$ to $76.968\%$) and
reduces the generalization gap (from $2.54\%$ to $1.5\%$). However, the impact of
combining PGT with label smoothing can vary depending on the value of the hyperparameter
$\alpha$. With a PGT hyperparameter value of $\alpha=0.05$, we notice the greatest
performance improvement, $1.246\%$ over the step scheduler test baseline and $0.656\%$
over the cosine scheduler test baseline. Adding label smoothing to PGT ($\alpha=0.05$)
hurts performance even though it reduces the generalization gap.

@@

\newpage

@@

\newpage \subsection{Grid search for different values of $\alpha$}

\newpage \subsection{Norm plots of weights (baseline (low batch size)) of non-BN
ResNet-18}
\begin{figure}[h] \centering \includegraphics[width=0.55\textwidth]{baseline-w_norm}
\caption{ Method: baseline, batch size = 256. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. } \end{figure}

\newpage \subsection{Norm plots of weights (baseline (high batch size)) of non-BN
ResNet-18}
\begin{figure}[h] \centering
\includegraphics[width=0.55\textwidth]{baseline_high_bs-w_norm} \caption{ Method:
baseline (high batch size), batch size = 1024. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. } \end{figure}

\newpage \subsection{Norm plots of weights (PGT) of non-BN ResNet-18}
\begin{figure}[h] \centering \includegraphics[width=0.55\textwidth]{pgt-w_norm}
\caption{ Method: PGT, batch size = 256. This is a plot depicting the iteration-wise
evolution of the norm of each filter in each layer. Each subplot corresponds to a layer
of the unnormalized ResNet-18. Above each subplot, the layer indices are shown. Each
filter in each subplot is shown with a separate and randomly determined colour. }
\end{figure}

\newpage \subsection{Norm plots of weights (AGC + PGT) of non-BN ResNet-18}
\begin{figure}[h] \centering \includegraphics[width=0.55\textwidth]{agc_pgt-w_norm}
\caption{ Method: AGC + PGT, batch size = 256. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. } \end{figure}

\newpage \subsection{ResNet-18 Layer Index Diagram}

>>>>

\begin{figure}[!t]
\centering
\begin{subfigure}{.5\columnwidth}
\centering
\includegraphics[width=0.9\columnwidth]{acc_vs_epoch_r18}
\caption{Plot of training and test accuracies (ResNet-18)}
\end{subfigure}%
\begin{subfigure}{.5\columnwidth}
\centering
\includegraphics[width=0.9\columnwidth]{acc_vs_epoch_r50}
\caption{Plot of training and test accuracies (ResNet-50)}
\end{subfigure}
\caption{ Log-linear plots of training and test accuracies and comparison with baseline
of batch-normalized variants: \textbf{(a)} ResNet-18 ($\alpha=0.25$), \textbf{(b)}
ResNet-50 ($\alpha=0.3$). For better visualization of accuracies obtained in the final
iterations, we plot both axes in log scale. PowerGrad Transform improves upon the
baseline in all cases. The corresponding accuracies are mentioned in table
\ref{tab:imagenet_table}. }
\label{fig:metrics}
\end{figure}

@@

In our experiments with Squeeze-and-Excitation variant of ResNet-50 i.e.
SE-ResNet-50\cite{hu2018squeeze}, we observe significant improvements; a $0.97\%$ boost
in training accuracy and a $0.734\%$ increase in test accuracy. These improvements are
obtained with the number of epochs being kept the same as that of the baseline model.
Alternatively, practioners can choose to employ PGT and have the networks converge to
accuracies similar to that of the baseline with a much lower epoch budget. We explore
this in detail in section \ref{sec:speed}. Therefore, PGT also speeds up training. In
our experiments with ResNet-50, we find an increase of $0.5\%$ and $0.656\%$ performance
enhancement over the baseline while using the cosine scheduler. The corresponding
improvement over the step scheduler baseline for ResNet-50 is $0.57\%$ and $0.524\%$
respectively for training and testing. ResNet-101 sees a higher improvement in training
fit, $0.81\%$ to be exact, while the improvement over the test set is $0.362\%$. Smaller
networks such as SE-ResNet-18 and ResNet-18 sees accuracy boosts which are smaller but
nevertheless positive. Therefore we conclude that with PGT, the consistent improvements
in training accuracies across all cases is because the networks train better and arrive
at better optimas during convergence. Per epoch training and test accuracy plots of
ResNet-18 and ResNet-50 (both with and without PGT) are shown in Fig.
\ref{fig:metrics}(a, b). For better visualization of the top end of the plots, we scale
both axes in log scale. Our experiments with batch-normalized networks indicate that
PowerGrad Transform speeds up training and allows the network to learn better
representations\footnote{Reproducible code, training recipes for the experiments,
pretrained checkpoints and training logs are provided at: \\
\url{ https://github.com/bishshoy/power-grad-transform }. Code is adapted from
\cite{rw2019timm}.}.

@@

\begin{table}[!t]
\centering
\caption{ Results for non-normalized ResNets on ImageNet-1K. Each experiment is
conducted with a 100 epoch budget using the cosine decay scheduler. Best training and
test accuracies are highlighted in red and blue respectively. Top differences in
training and test accuracies are marked in yellow. The baselines in each segment is
taken as the reference point for all `Diff' accuracies in that segment. The first three
segments highlight comparison for experiments where the methods are applied
individually, while the last segment denotes performance metrics for combinations of
different methods (GC + PGT, AGC + PGT). We see that at any batch size, with PGT, the
performance of non-normalized version of ResNet-18 improves significantly over the
baseline owing to the improvements in training as observed in Fig. \ref{fig:agc_pgt}. }
\label{tab:wobn_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Batch Size}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PowerGrad} & \textbf{\hspace{-0.65cm} Train}
& \textbf{Train} & \textbf{\hspace{-0.65cm} Test} & \textbf{Test} \\
& & & \textbf{Transform ($\alpha$)} & \textbf{Accuracy (\%)} & \textbf{Diff (\%)} &
\textbf{Accuracy (\%)} & \textbf{Diff (\%)} \\
\midrule
ResNet-18 (non-BN) & 1024 & Baseline & - & 66.27 & - & 64.816 & - \\
& 1024 & PGT & 0.92 & \textcolor{red}{\textbf{66.62}} &
\textcolor{olive}{\textbf{+0.35}} & \textcolor{blue}{\textbf{65.498}} &
\textcolor{olive}{\textbf{+0.682}} \\
\cmidrule{2-8}
& 512 & Baseline & - & 68.02 & - & 66.552 & - \\
& 512 & PGT & 0.25 & \textcolor{red}{\textbf{69.5}} & \textcolor{olive}{\textbf{+1.48}} &
\textcolor{blue}{\textbf{67.236}} & \textcolor{olive}{\textbf{+0.684}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC & - & 69.04 & +0.18 & 67.064 & +0.268 \\
& 256 & AGC & - & 69.06 & +0.2 & 67.298 & +0.502 \\
& 256 & PGT & 0.25 & \textcolor{red}{\textbf{69.97}} & \textcolor{olive}{\textbf{+1.11}} &
\textcolor{blue}{\textbf{67.814}} & \textcolor{olive}{\textbf{+1.018}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC+PGT & 0.25 & 68.67 & -0.19 & 67.088 & +0.292 \\
& 256 & AGC+PGT & 0.25 & \textcolor{red}{\textbf{70.92}} & \textcolor{olive}{\textbf{+2.06}} &
\textcolor{blue}{\textbf{68.856}} & \textcolor{olive}{\textbf{+2.06}} \\
\end{tabular}
\end{table}

We provide performance metrics on the test set of ImageNet-1K for different batch sizes
for the non-BN ResNet-18 and also experiment with different methods (GC, AGC, PGT along
with their combinations). Table \ref{tab:wobn_table} shows the findings. To begin, we
use a batch size of 1024. The baseline performance for this batch size is drastically
inferior to baselines for other batch sizes. PGT helps regain some of the lost
performance by $0.682\%$ ($65.498\%$ vs. $64.816\%$). This illustrates that in
unnormalized networks, accuracy suffers significantly when batch sizes are large.
Additionally, the training process is often unstable at this batch size, resulting in
filters and sometimes entire layers frequently zeroing out (as seen in \ref{sec:Empi}).
At a batch size of 512, invoking PGT improves the training accuracy baseline by $1.48\%$
and the test accuracy baseline by $0.684\%$, while at a batch size of 256, the
improvement in training and test accuracies are $1.11\%$ and $1.018\%$ respectively. In
comparison, the test accuracy improvements obtained by GC and AGC at batch size of 256
is much less, at $0.27\%$ and $0.5\%$, respectively.

@@

We conduct an ablation study to investigate the effects of PowerGrad Transform for
different values of the hyperparameter ($\alpha$). Grid plots are made available in the
supplementary section. We have previously demonstrated from the experiments depicted in
section \ref{sec:Expe}, that PowerGrad Transform causes the network to arrive at a
better optima and fit the training dataset better than the baseline. In this ablation
study, we use the ResNet-50 architecture and we combine our proposed method with other
regularization techniques such as label smoothing and report our findings in table
\ref{tab:ablation_table}. We also investigate the two optimal peaks of the
hyperparameter ($\alpha$) obtained from ResNet-50's grid search ($\alpha=0.05$ and
$\alpha=0.3$). First, we examine the effect of PGT on the step scheduler baseline in
order to later compare it to the cosine scheduler baseline. PGT improves upon the step
scheduler baseline (test set) by a substantial margin with $0.524\%$ ($76.494\%$ as
opposed to $75.97\%$). Introducing the cosine scheduler yields a $0.59\%$ improvement
($76.56\%$ vs. $75.97\%$) over the step scheduler. After introducing label smoothing,
the test accuracy relative to the cosine scheduler baseline increases by only $0.138\%$
(from $76.56\%$ to $76.698\%$). However, introducing PGT with $\alpha=0.3$ alone
(without label smoothing) improves the cosine scheduler baseline by $0.326\%$
($76.886\%$ vs. $76.56\%$). Combining PGT ($\alpha=0.3$) with label smoothing improves
the performance on the test set further by $0.408\%$ (from $76.56\%$ to $76.968\%$) and
reduces the generalization gap (from $2.54\%$ to $1.5\%$). However, the impact of
combining PGT with label smoothing can vary depending on the value of the hyperparameter
$\alpha$. With a PGT hyperparameter value of $\alpha=0.05$, we notice the greatest
performance improvement, $1.246\%$ over the step scheduler test baseline and $0.656\%$
over the cosine scheduler test baseline. Adding label smoothing to PGT ($\alpha=0.05$)
hurts performance even though it reduces the generalization gap.

@@

@@

\subsection{Grid search for different values of $\alpha$}

\begin{figure}[h]
\centering
\includegraphics[width=0.5\columnwidth]{r18_grid_search}
\caption{ Grid search for various values of $\alpha$ for ResNet-18 architecture with the
ImageNet dataset. Each datapoint is collected over a 50 epoch budget, with other
hyperparameters kept the same as mentioned in section \ref{sec:bn}. }
\label{fig:grid_searcha}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=0.5\columnwidth]{r50_grid_search}
\caption{ Grid search for various values of $\alpha$ for ResNet-50 architecture with the
ImageNet dataset. Each datapoint is collected over a 50 epoch budget, with other
hyperparameters kept the same as mentioned in section \ref{sec:bn}. }
\label{fig:grid_searchb}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=0.5\columnwidth]{cifar10_grid_search}
\caption{\\ResNet-20 CIFAR-10}
\caption{ Grid search for various values of $\alpha$ for ResNet-20 architecture with the
CIFAR-10 dataset. Each datapoint is collected over a 360 epoch budget. }
\label{fig:grid_searchc}
\end{figure}

\subsection{Norm plots of weights (baseline (low batch size)) of non-BN
ResNet-18}
\begin{figure}[h] \centering \includegraphics[width=0.55\textwidth]{baseline-w_norm}
\caption{ Method: baseline, batch size = 256. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. } \end{figure}

\subsection{Norm plots of weights (baseline (high batch size)) of non-BN
ResNet-18}
\begin{figure}[h] \centering
\includegraphics[width=0.55\textwidth]{baseline_high_bs-w_norm} \caption{ Method:
baseline (high batch size), batch size = 1024. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. } \end{figure}

\subsection{Norm plots of weights (PGT) of non-BN ResNet-18}
\begin{figure}[h] \centering \includegraphics[width=0.55\textwidth]{pgt-w_norm}
\caption{ Method: PGT, batch size = 256. This is a plot depicting the iteration-wise
evolution of the norm of each filter in each layer. Each subplot corresponds to a layer
of the unnormalized ResNet-18. Above each subplot, the layer indices are shown. Each
filter in each subplot is shown with a separate and randomly determined colour. }
\end{figure}

\subsection{Norm plots of weights (AGC + PGT) of non-BN ResNet-18}
\begin{figure}[h] \centering \includegraphics[width=0.55\textwidth]{agc_pgt-w_norm}
\caption{ Method: AGC + PGT, batch size = 256. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. } \end{figure}

\subsection{ResNet-18 Layer Index Diagram}
